---
title: "Tutorial 2 - Statistics"
author: "Almog Angel"
date: "24/01/2024"
output: html_document
---

## Corona cabinet challenge

```{r}
library(tidyverse)
# Also install the package "plyr"

vac_by_city <- read_csv("vaccinated_city_table_ver_00246.csv")
vac_by_city
```

## Orginaze data

- Transform data to long data format for using maximum of Tidyverse's power!
```{r}
vac_by_city <- vac_by_city %>% 
  pivot_longer(cols = -c(CityName, CityCode, Date), names_to = "age_group", values_to = "n_vac") %>%
  filter(Date == "2021-10-26") %>%
  separate(age_group, into = c("dose","dose2", "age_group"), sep = "_") %>%
  select(-dose2) %>%
  filter(dose == "third")
vac_by_city
```

- Convert n_vac column to numeric
```{r}
# We will assume that <15 equal 10 people
vac_by_city <- vac_by_city %>% 
  mutate(n_vac = ifelse(n_vac == "<15", 10, n_vac)) %>% 
  mutate(n_vac = as.numeric(n_vac))

vac_by_city
```

- Calculate the total number of people vaccinated per city
```{r}
vac_by_city <- vac_by_city %>% 
  group_by(CityName, CityCode, Date) %>% 
  summarise(total_vac_per_city = sum(n_vac)) %>% 
  ungroup()

vac_by_city
```


- Load data about the population of each city
```{r}
city_code2population <- read_csv("city_code2population.csv")
city_code2population
```

- Assign data about population of cities by city_code
```{r}
vac_by_city <- vac_by_city %>% 
  left_join(city_code2population, by = c("CityCode" = "city_code")) %>% 
  rename(city_population = total)

vac_by_city
```

- Calculate the relative vaccination per city
```{r}
vac_by_city <- vac_by_city %>% 
  mutate(relative_vac = round(total_vac_per_city/city_population*100, 2))

vac_by_city
```

- Check for NAs
```{r}
vac_by_city[is.na(vac_by_city$relative_vac),]
```

- Drop NAs
```{r}
vac_by_city <- vac_by_city %>% 
  drop_na()

vac_by_city
```

- Visualize the data (does something looks wrong?)
```{r}
vac_by_city %>% 
  ggplot(., aes(relative_vac)) +
  geom_histogram(position = 'identity', bins = 50)
```


## Run statistical test

- Use most updated data
```{r}
vac_by_city <- vac_by_city %>% 
  filter(Date == "2021-10-26")

vac_by_city
```


- Sample 7 cities per group
```{r}
center_cities <- c("Tel Aviv-Yafo", "Ramat Gan", "Giv'atayim", "Holon", "Bat Yam", "Rehovot", "Ramat HaSharon", "Herzliya", "Ra'anana", "Kiryat Ono", "Yavne", "Rishon-LeZion", "Netanya", "Kfar Saba", "Hod HaSharon", "Petah Tikva", "Or Yehuda", "Yehud", "Ramla", "Lod", "Ness Ziona", "Modi'in-Maccabim-Re'ut", "Kiryat Motzkin", "Giv'at Shmuel", "Rosh HaAyin", "Kiryat Tiv'on", "Shoham", "Savyon", "Kfar Yona", "Mazkeret Batya")

center <- c(5000, 8600, 6300, 6600, 6200, 8400, 2650, 6400, 8700, 2620, 2660, 8300, 7400, 6900, 9700, 7900, 2400, 9400, 8500, 7000, 7200, 1200, 8200, 681, 2640, 2300, 1304, 587, 168, 28)

perip_cities <- c("Ofakim", "Eilat", "Ashdod", "Ashkelon", "Be'er Sheva", "Beit She'an", "Dimona", "Yeruham", "Nahariya", "Arad", "Kiryat Gat", "Kiryat Shmona", "Katzrin", "Shlomi", "Afula", "Hatzor HaGlilit", "Tiberias", "Netivot", "Nesher", "Kiryat Malachi", "Sderot", "Kiryat Ekron", "Rosh Pina", "Karmiel", "Ma'alot-Tarshiha", "Nof HaGalil", "Acre", "Hadera", "Shefar'am", "Tamra")

periphery <- c(31, 2600, 70, 7100, 9000, 9200, 2200, 831, 9100, 2560, 2630, 2800, 4100, 812, 7700, 2034, 6700, 246, 2500, 1034, 1031, 469, 26, 1139, 1063, 1061, 7600, 6500, 8800, 8900)


cent_vs_perip <- vac_by_city %>% 
  ungroup() %>% 
  filter(CityCode %in% c(center, periphery)) %>% 
  mutate(CityName = plyr::mapvalues(CityCode, from = c(center, periphery), to = c(center_cities, perip_cities)))

unique(cent_vs_perip$CityName)


```


- Add a column that say whether the city is in the periphery or center
```{r}
cent_vs_perip <- cent_vs_perip %>% 
  mutate(location = ifelse(CityCode %in% periphery, "Periphery", "Center"))

cent_vs_perip
```

- Visualize the data with boxplot
```{r}
ggplot(cent_vs_perip, aes(x=location, y=relative_vac)) +
  geom_boxplot(aes(fill=location)) +
  geom_point() +
  geom_text(aes(label=CityName), hjust=0.5, vjust=-1)

```

- Visualize the data with histograms
```{r}
ggplot(cent_vs_perip, aes(x=relative_vac, fill = location)) +
  geom_histogram(position = 'identity', bins = 50, alpha = 0.5)
```


** Test for normality: ***
```{r}
shapiro.test(cent_vs_perip[cent_vs_perip$location == "Center",]$relative_vac)
shapiro.test(cent_vs_perip[cent_vs_perip$location == "Periphery",]$relative_vac)
```



## ToDo:
- Define the null and alternative hypothesis
- Run a t-test using the t.test() function 
- Report the t- and p-values
- What is your conclusion?

```{r}
# Run t-test here:
cent_vs_perip$location <- factor(cent_vs_perip$location, levels = c("Center", "Periphery"))
t.test(cent_vs_perip$relative_vac ~ cent_vs_perip$location, alternative = "greater")

```