---
title: "Tutorial 3 - Sequence Alignment"
author: "Almog Angel"
date: "31/1/2024"
output: html_document
---

## Class exercise

- Load packages and set working directory
```{r}
library(Biostrings)
library(msa)
library(msaR)

setwd("~/Documents/intro_to_bioinformatics/winter2324/tutorial_3/")
```

- Load Wuhan's covid genome
```{r}
wuhan.covid.genome <- readDNAStringSet("wuhan_covid_genome.fasta")
wuhan.covid.genome
```

- Find how many start codons ("ATG") exist in the genome
```{r}
vmatchPattern("ATG", wuhan.covid.genome)  # 725
# However, the virus have only about 15 genes
```

- Find the reverse complement sequence of the first 10bp
```{r}
first10bp <- wuhan.covid.genome[[1]][1:10]
first10bp
reverseComplement(first10bp)
```

- Find the G/C content of the COVID-19 genome
```{r}
letterFrequency(wuhan.covid.genome, letters = "GC", as.prob = TRUE)
```

- Load the protein sequence of the spike protein from wuhan_spike.fasta
```{r}
wuhan.covid.spike.aa <- readAAStringSet("wuhan_spike.fasta")
wuhan.covid.spike.aa
```

- Extract the first 5 amino acids from the protein sequence
```{r}
first_5aa <- wuhan.covid.spike.aa[[1]][1:5]
first_5aa
```

## Task 1 - Extract the spike DNA sequence

- Look at the genetic code (from DNA to AA)
```{r}
GENETIC_CODE
```

- In some cases, the same amino acid can be originated from different codons
- Here are 5 possible DNA sequences that can be translate to the same amino acid sequence
```{r}
possible_dna_seq <- DNAStringSet(c("ATGTTTGTGTTCCTG",
                      "ATGTTTGTATTCCTA",
                      "ATGTTTGTTTTTCTT",
                      "ATGTTCGTTTTTCTC",
                      "ATGTTCGTTTTCCTT"))
Biostrings::translate(possible_dna_seq)
```

- However, only one sequence completely match the beginning of the COVID-19 spike sequences
- Find out which sequence match by using the function pairwiseAlignment()
- First, read the documentation by running ?pairwiseAlignment
- Save your results to a variable named "alignment.out"

```{r}
??pairwiseAlignment

# Write your code here:
alignment.out <- pairwiseAlignment(pattern = possible_dna_seq, subject = wuhan.covid.genome, type = "local")
```

- Look at the results
```{r}
alignedPattern(alignment.out)
```

- Which sequence have complete match? assign the sequence number to "n"
- Find the position where the spike gene start
```{r}
# assign the sequence number to "n"
n <- 3
alignment.out[n] 
```

- Assign the position to "start"
```{r}
start <- 21563 # insert start position
```

- Find the position where the gene ends and assign to "end" (Hint: used the width() function)
- Extract the spike gene sequence and assign to "covid.spike.dna"
```{r}
end <- start + width(wuhan.covid.spike.aa)*3 + 2 # +2 for the stop codon
#start:(start+3)
covid.spike.dna <- wuhan.covid.genome[[1]][start:end]
covid.spike.dna
```


## Task 2 - Multiple Sequence Alignments

- Read "spikes.fasta" into "all.spikes"
- This file contain DNA sequence of spike proteins from humans (Hu), bats (Bt), civet (Cv), camel (Cm) and pangolin (Pn)
```{r}
# Compare covid19 to other viruses
all.spikes <- readDNAStringSet("spikes.fasta")
names(all.spikes)
```

- Change to more friendly names
```{r}
names(all.spikes) <- c("Pn_SARS-CoV-2", "Bt_SARS-CoV-2", "Bt_SARS-CoV",
                         "Cv_SARS-CoV", "Hu_SARS-CoV", "Bt_MERS-CoV",
                         "Cm_MERS-CoV", "Hu_MERS-CoV")
all.spikes
```

- Add the human SARS-CoV-2 spike sequence we extracted in task 1
```{r}
all.spikes[[9]] <- covid.spike.dna
names(all.spikes)[9] <- "Hu_SARS-CoV-2"
all.spikes
```


- Run MSA for all sequences in "all.spikes" (might take a 1-2 minutes to finish)
- Assign the output to "all.spikes.msa"
```{r}
all.spikes.msa <- msa(inputSeqs = all.spikes, type = "dna")
all.spikes.msa
```

- Visualize results
```{r}
msaR(DNAMultipleAlignment(all.spikes.msa), colorscheme = "taylor")
```


## Task 3 - Construct a phylogenetic tree

- Now we will construct a phylogenetic tree form the MSA results

- Run this code to convert msa object to seqinr object
```{r}
covid.spikes.msa.seqinr <- msaConvert(all.spikes.msa, type="seqinr::alignment")
```

- Run this code to calculate a distance matrix of all sequences
```{r}
covid.spikes.msa.distmat <- seqinr::dist.alignment(covid.spikes.msa.seqinr, matrix = "similarity")
covid.spikes.msa.distmat
```

- Finally we can look at the phylogenetic tree (dendrogram) assembled by hierarchical clustering
```{r}
covid.spikes.msa.clusters <- hclust(covid.spikes.msa.distmat, method = "average")
plot(covid.spikes.msa.clusters, hang = -1, main = "Spikes phylogenetic three", xlab = "Organism", ylab  = "Distance")
```


- Explain what you see in the phylogenetic three